R course Part 1 - Pseudo Facebook
========================================================

[Dataset](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/pseudo_facebook.tsv) taken from Udacity R course lesson 3

***

### What to Do First?
Notes: EDA is an opportunity to let data surprise you and learn from it. EDA will help you understand your variables and that will need you to look at data and make some good visualizations. Also, it is very important that you think back about what question you want to answer with your data.

***

### Pseudo-Facebook User Data
Loading the data **pseudo_facebook.tsv** from the directory in which we stored it and using the right delimiter (the extension *\*.tsv* means *tab separated file*).

```{r Pseudo-Facebook User Data}
# Check current path
path <- getwd()

# Set path to dataset
path <- setwd('~/github/DataAnalyst/Rbasics_Appendix/pseudoFacebookAnalysis/')

file <- 'pseudo_facebook.tsv'

# Read file
pf <- read.csv(file, sep='\t') 
# can use read.delim(file) as its default value for separator is \t
```

***

### Histogram of Users' Birthdays
Plot a histogram of the day of birthday of users using the qplot function from ggplot2 library.
>Note: We can change the theme in which we plot the histogram using the ggthemes library. We can set the theme with theme_set() function. Other ways to set themes and more themes available at [ggthemes github](https://github.com/jrnold/ggthemes#geoms).

```{r Histogram of Users\' Birthdays}
# Load ggplot2 library
#install.packages('ggplot2')
library(ggplot2)

# Load ggthemes
#install.packages('ggthemes', dependencies = TRUE)
library(ggthemes)

# Set the theme to theme_minimal and font size 24
theme_set(theme_minimal(8))

# Show the columns of the dataset
names(pf)

# Plot a histogram with default configuration
qplot(data = pf, x = dob_day)

```

After plotting the data we noticed that looking at it is not very easy, so we need to improve the visualization of this histogram. We can show all the days in the x axis using ```scale_x_continuos()``` function.

```{r}
# Plot a histogram with all days of a month shown as x ticks
# qplot(data = pf, x = dob_day) + scale_x_continuous(breaks=1:31)

# Also plotting the same as above but using ggplot instead of qplot function
ggplot(aes(x = dob_day), data = pf) +
  geom_histogram(binwidth = 0.5 ) +
  scale_x_continuous(breaks = 1:31)

```

***

#### What are some things that you notice about this histogram?
Notice that in this plot the first day has a lot more users, which would mean that more people is borned on the first day of a month or that there is a default day of birthday when signing up. On the other hand, we can see that 31 is the lowest bar and it can be caused due to half of the months have only 30 days.

***

### Moira's Investigation 
We are going to reproduce a study done by a data scientist at Facebook, Moira Burke. 

***

### Estimating Your Audience Size
Moira and her team analyzed users publications and audience and found people tend to understimate their audience size. You can find more in their paper about ["Quantifying the Invisible Audience in Social Networks"](http://hci.stanford.edu/publications/2013/invisibleaudience/invisibleaudience.pdf).

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
A trip to Japan

#### How many of your friends do you think saw that post?
Around 200 of them

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
I think only 20% of my friends saw it.

***

### Perceived Audience Size
As Moira and her team found that usually the audience was underestimated by a quarter of the size, we are going to plot histograms to understand what people guessed about their audience size.

***
### Faceting
Before we plotted all our days of the year/month in one single plot. We would like to separate that plot in histograms per month and look at them. We can use faceting to do this. There are two options for faceting, here we will use ```facet_wrap()``` as it is easier to use when dealing with one variable. However, when dealing with more than one variable it might be better to use ```facet_grid()```, to learn more about how to use it read this [documentation](http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/).

```{r Faceting}

# Plot with month as facet (3 columns, meaning 3 months per row)
ggplot(aes(x = dob_day), data = pf) +
      geom_histogram(binwidth = 0.5) +
      scale_x_continuous(breaks = 1:31) +
      facet_wrap(~dob_month, ncol = 2)

```

#### Let’s take another look at our plot. What stands out to you here?
Three main observations:
* January first is the highest bar and maybe it is not normal as all other days in the year are not even a quarter of it. 
* Some months wihtout 31 day on them have no birthdays which is a good indication 
* February 29 has the smallest amount a day can have birthdays on and that can be related to the calendar in which only every 4 years we have February 29.

***

### Be Skeptical - Outliers and Anomalies
The approach to deal with outliers depends on what your analysis is looking to answer since outliers can have many causes, for example:

* Accurate data about an extreme case
* Bad data
* Limitations of data

One of the approaches used is to replace extreme values with normal values or mean values.

***

### Moira's Outlier
In this section, we will be checking an outlier in the data of the audience estimation analysis.

#### Which case do you think applies to Moira’s outlier?
Bad data of an extreme case since you cannot have 1,000,000 friends on facebook and the data showed 10,000,000 friends watched a post. 
***

### Friend Count


#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(data = pf, x = friend_count)
```

#### How is this plot similar to Moira's first plot?
It has some values that are very large and do not let us appreciate all the data properly.

***

### Limiting the Axes
In order for us to better see and understand the data, we want to limit the axes. We have two options:
* ```xlim``` parameter inside ```qplot()```
* ```scale_x_continuous(limits = c(low,high))`` function added to ```qplot()```

```{r Limiting the Axes}

# option 1, use xlim
qplot(data = pf, x = friend_count, xlim = c(0,1000))

#option 2, use x_scale_continuous
qplot(data = pf, x = friend_count) + scale_x_continuous(limits=c(0,1000))
```

### Exploring with Bin Width
In the plot above, we can see that we have bins in ranges, it will be good to change the ```binwidth``` to explore our data and find how it all distributes with different bin ranges. It will be a good idea to show more ticks by using the ```breaks``` argument in ```scale_x_continuous()```.

***

### Adjusting the Bin Width
```{r Adjusting Bins}
qplot(data = pf, x = friend_count, binwidth=25) +
  scale_x_continuous(limits = c(0,1000), breaks = seq(0,1000,50))
```

### Faceting Friend Count
Now, it will be interesting to split the histogram by gender to understand what males or females think about their audience.
```{r Faceting Friend Count}
# Facet the histogram by gender
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)
```

***

### Omitting NA Values
We noticed that the plot above had a third plot with title NA. This means some users did not define their gender. To remove the NA data we can ```subset``` our dataset using ```!is.na()``` with the ```gender``` column.

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)
```
**Note:** We can ommit NA values using ```na.omit(DATASET)```, but it will delete all rows where NA values are found. We might not want this for this situation as some other columns might have NA values, deleting rows in which we have valid gender values. 
***

### Statistics 'by' Gender
After looking at the two histograms, we noticed that males seem to have more friends, so we want to answer: Who has more friends? Whom do we have more entries?

```{r Statistics \'by\' Gender}
# count records per gender
table(pf$gender)

by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Women on average has more friends with a mean of 242 friends.

#### What's the difference between the median friend count for women and men?
The difference between women's and men's median is 22, which is smaller than the difference in means.

#### Why would the median be a better measure than the mean?
The median is a better measure in this ocassion since the histograms are skewed to the right.

***

### Tenure
Now we want to create a histogram of tenure, and at the same time improve our knowledge of histogram formats, so we will add color to it.

```{r Tenure}
qplot(x = tenure, data = pf, binwidth = 30,
        color = I('black'), fill = I('#099DD9'))
```

**Note:** An alternative way to plot the same histogram using ```ggplot()``` is:
```
ggplot(aes(x = tenure), data = pf) +
  geom_histogram(binwidth = 30, color = 'black', fill = '#099DD9')
```

***

#### How would you create a histogram of tenure by year?

In the previous histogram tenure is expressed on days, so we would like to show it in years

```{r Tenure Histogram by Year}
qplot(x = tenure/365, data = pf, binwidth = 0.5,
        color = I('black'), fill = I('#099DD9')) +
  scale_x_continuous(breaks = seq(1,7,1), limits = c(0,7.5))
```

Notice that the binwidth size reduced as we divided tenure by 365. Then our bin width should decrease as well as the values of tenure did.
Also, we decided to cut the x axis to show only values until 7.5 as after this value it seems we almost have 0 users with tenure > 7.5 years.

***

### Labeling Plots
Once we have our tenure plot by years, we can see that the xlabel and y label are not very intuitive, so we need to change them for something more meaningful.

```{r Labeling Plots}
qplot(x = tenure/365, data = pf, 
        xlab = 'Number of Years using Facebook',
        ylab = 'Number of Users',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1,7,1), limits = c(0,7.5))

```

**Note:** A way to do the same as above using ```ggplot``` is:
```
ggplot(aes(x = tenure / 365), data = pf) +
  geom_histogram(color = 'black', fill = '#F79420') +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7)) +
  xlab('Number of years using Facebook') +
  ylab('Number of users in sample')
```
***

### User Ages
Now, we want to know what is the distribution of ages of Facebook users.

```{r User Ages}
qplot(x = age, data = pf, binwidth = 1,
      xlab = 'Users Age',
      ylab = 'Number of Users',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(10,115,5), limits = c(0,115))
```

#### What do you notice?
The highest number of users is from 12 to 35 years old. We also have some fliers at 103 and 107 years old. It might be good to investigate the fliers as it seems very odd that old people have such spike, maybe it is companies who have years on the market.

***

### Transforming Data
If we encounter **overdispersed** data, we can apply transformations to our data. A log transformation is very common as the logarithm is used to represent big manitudes into small order magnitudes. 
Another useful transformation is the square root transform.
We will apply these transformations to our age and friends count data to see the outcomes.

**Note:** Overdispersed data means that some of our values have a frequency of ocurrance greater by a big amount (around 100 times) compared to other values.

```{r Transforming Data}

summary(pf$friend_count)

# we add +1 to avoid log 0 which is infinite
summary(log10(pf$friend_count + 1)) 
summary(sqrt(pf$friend_count))

#install.packages('gridExtra')
library(gridExtra)

f1 <- ggplot(aes(x = friend_count), data = pf) +
      geom_histogram(binwidth = 5, color = 'black', fill = '#F79420') + 
      xlab('Number of Friends') +
      ylab('Frequency') +
    scale_x_continuous(breaks = seq(0,300,50), limits = c(0,300))

f2 <- ggplot(aes(x = log10(friend_count + 1)), data = pf) +
      geom_histogram(binwidth = .2, color = 'black', fill = '#F79420') + 
      xlab('Number of Friends') +
      ylab('Frequency') +
    scale_x_continuous(breaks = seq(0,10,1), limits = c(0,10))

f3 <- ggplot(aes(x = sqrt(friend_count)), data = pf) +
      geom_histogram(binwidth = 0.75, color = 'black', fill = '#F79420') + 
      xlab('Number of Friends') +
      ylab('Frequency') +
    scale_x_continuous(breaks = seq(0,60,10), limits = c(0,60))


grid.arrange(f1, f2, f3, ncol=1)
```

***
There is another syntax we can use for plotting multiple plots using ```gridExtra``` library. We use it below to plot the age data.

```{r}

summary(pf$age)
summary(log10(pf$age + 1))
summary(sqrt(pf$age))


p1 <- ggplot(aes(x = age), data = pf) +
    geom_histogram(color = 'black', fill = '#F79420') + 
    scale_x_continuous(breaks = seq(10,115,5), limits = c(0,115)) +
    #xlab('Users Age') +
    ylab('Number of Users')

p2 <- p1 + scale_x_log10()

p3 <- p1 + scale_x_sqrt()

grid.arrange(p1, p2, p3, ncol = 1)
```

***
**Note:** We can also use qplot.
```
p1 <- qplot()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()

grid.arrange(p1,p2,p3, ncol = 1)

```

From the plots above we can see that the log scale is a good fit since it fits our data making it a distribution close to normal.

### Add a Scaling Layer
If you look at the plots above you can see the scales when using ```log10()``` inside ggplot/qplot is different that when we used ```scale_x_log10()```. Let's look this more in detail using the ```friend_count``` column. 

```{r Add a Scaling Layer}
p1 <- qplot(x = log10(friend_count + 1), data = pf)
p2 <- qplot(x = friend_count, data = pf ) + scale_x_log10()
grid.arrange(p1, p2, ncol = 2)

```

The difference on the plots:
* ```log10()``` inside qplot/ggplot x value uses the scale of the value we calculate from the column data.
* ```scale_x_log10()``` uses the actual scale of the column we use in x.

***


### Frequency Polygons
In order to compare the two histograms we can make a frequency polygon instead of calling ```facet_wrap()```. We need to specify in ```qplot()``` the parameter ```geom = 'freqpoly``` and to identify the graphs better we can assign a color to each of them using the parameter ```color = gender```

```{r Frequency Polygons}
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Frequency',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(lim = c(0,1000), breaks = seq(0,1000,50))
```

Another improvement we can make is to show the proportion of users with a specific friend count rather than the number of users. In order to do such thing, we need to add a parameter ```y``` in qplot and define ti as the proportion value.

```{r}
qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)), 
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that Friend Count',
      binwidth = 10, geom = 'freqpoly', color = gender) + 
  scale_x_continuous(breaks = seq(0,300,50), lim = c(0,300))
```

**Note:** If you want to reproduce the same plot using ggplot here is how:
```
ggplot(aes(x = friend_count, y = ..count../sum(..count..)),
       data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') +
  ylab('Proportion of users with that friend count')
```

***

### Likes on the Web
Since we got the proportion of friends per gender, we want to know the likes they got from web pages.

```{r Likes on the Web}
ggplot(aes(x = www_likes ),
       data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender)) +
  scale_x_continuous() +
  scale_x_log10() +
  xlab('Web Likes') +
  ylab('Proportion of Web Likes Per Gender')

```

We can tell that males start with more likes, but eventually females have more likes. This doesn't help us answer our original question of who has more likes?

**Note:**  There is another alternative to plot the data, using ```qplot()```

```{r}
qplot(x = www_likes, data = subset(pf, !is.na(gender)), geom = 'freqpoly',
      color = gender) +
  scale_x_continuous() + scale_x_log10()
```

In order to answer our question, let's find the number of likes counts for males and females.

```{r}
by(pf$www_likes, pf$gender, sum)

```

From our calculation above we can see that females have more likes. This matches with the females having more users with more likes in the histogram. We can now confirm that the females actually have more likes.

***

### Box Plots
Let's look at the friend counts' distributions by gender. 

```{r Box Plots}
qplot(x = gender, y = friend_count, data = subset(pf,!is.na(gender)), 
      geom = 'boxplot')
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.
We have many outliers, so we are going to limit the friend count from 0 to 1000.

**Note:** An outlier is a value 1.5 times greater than the IQR.

```{r}
qplot(x = gender, y = friend_count, data = subset(pf,!is.na(gender)), 
      geom = 'boxplot', ylim = c(0,1000))
```


We used the parameter ```ylim``` to adjust the boxplot to the range 0,1000; however, this parameter removes all the points beyond 1000. This will give a different boxplot since those points won't be used in the calculation of the IQR.
We need to use a new function for the plot to have the true box plot: ```coord_cartesian()```

```{r}
qplot( x = gender, y = friend_count, data = subset(pf, !is.na(gender)),
       geom = 'boxplot') +
  coord_cartesian(ylim = c(0,1000))
```

Now, the boxplot with ```coord_cartesian()``` give us a different IQR which used all the values we have in our first plot. This is the boxplot we want to inspect and no the previous plot using ```ylim```. 

***

### Box Plots, Quartiles, and Friendships
It will be a good idea to show that the ``coord_cartesian()``` function matches the IQR from our friend count. We will need to zoom in the IQR are of our previous plot.

```{r Box Plots, Quartiles, and Friendships}
by(pf$friend_count, pf$gender, summary)

qplot(x = gender, y = friend_count, data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0,250))
```

From the summary, you can see that the females data matches the plot as the first quartile is 37 and the therd quartile is 244, those are very close values to our plot. The same happens for males, where the first quartile is 27 and the third quartile is 182, both values match our boxplot. 

#### On average, who initiated more friendships in our sample: men or women?

Using the same technique above let's explore who males or females inititated more friendships.

```{r Friend Requests by Gender}

by(pf$friendships_initiated, pf$gender, sum)

by(pf$friendships_initiated, pf$gender, summary)

qplot(x = gender, y = friendships_initiated, 
      data = subset(pf, !is.na(gender)), geom = 'boxplot') + 
  coord_cartesian(ylim = c(0,150))

```

We can see that men initiated more friendships, but on average women initiated more friendships. The median of the distributions also show that women started more friendships, 4 more frienships to be precise.

The boxplot supports our conclusion and it is very useful to comprehend the distribution of our data and to find which measure of center is better in our analysis.

***

### Getting Logical
As an extra note, there is another method to process our data and get the results we want. Here we will use the mobile_likes to determine how many users signed in using their mobile. We will need to binarize the mobile_likes column. We can do this using ```ifelse()``` in mobile_likes which if it is greater than 0 means the user signed in using his/her mobile. We can store this data in a new variable that we can use to perform an analysis on it.

```{r Getting Logical}
summary(pf$mobile_likes)

# Show the users that did mobile likes and the ones who didn't
summary(pf$mobile_likes > 0)

mobile_check_in <- NA
# Binarize the data
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
# Create a factor out of the binary data
pf$mobile_check_in <- factor(pf$mobile_check_in)
# Show a summary of the data
summary(pf$mobile_check_in)

# Calculate the percentage of mobile sign in users 
pct_mobile <- 100.0 * sum(pf$mobile_check_in == 1) / length(pf$mobile_check_in)
pct_mobile
```

From the analysis above we can use it to make a decision on if we should spend time in our mobile development or not.

The percentage of users using the mobile sign in is 64.59%. It might be worth it to spend time developing the mobile application.

***

### Analyzing One Variable
After going through all that exercise and some bonus exercises along the way, we can conclude that R basic skills to plot and summarize datasets are very useful skills when analyzing a single variable in our dataset.

We demosntrated the power of R when it comes to splitting data by categories using ```facet_wrap``` and we also show how R can summarize data in different ways using the function ```by()```. All in all, this introductory exercise to R is worth the effort to getting started with some R capablities when analyzing a single variable using different cathegories and when plotting data.
