R course Part 3 - Pseudo Facebook
========================================================

In part 3 of our pseudofacebook dataset analysis we will explore more than 2 variables at a time.

```{r Load Data}
# Check current path
path <- getwd()

# Set path to dataset
path <- setwd('~/github/DataAnalyst/Rbasics_Appendix/pseudoFacebookAnalysis/')

file <- 'pseudo_facebook.tsv'

pf <- read.csv(file, sep = '\t')

# load ggplot
library(ggplot2)

```

### Third Qualitative Variable

Let's start exploring our age distributions by gender, we can create a boxplot for each gender. We can identify the mean in the box plot using ```stat_summary(fun.y=mean, geom = 'point', shape=4)```, where the shape being 4 is the x symbol, and the geom = point marks the mean only as a point.

Another graph we can do is the age and gender count using the median as the measure of center.

```{r Third Qualitative Variable}
# Age by gender boxplots
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + 
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4)

# graph of friend count by gender
ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) +
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)
```

Let's now create a dataframe that groups our users by age and gender to obtain the mean of friend count, median of friend count and members of each group.

```{r}
library(dplyr)

# filter instead of using subset inside group_by()
pf_filter_gender <- filter(pf,!is.na(gender))

# group by age and gender
age_gender_group <- group_by(pf_filter_gender, age, gender)

# summarise the data in a dataframe
pf.fc_by_age_gender <- summarise(age_gender_group,
                                 mean_friend_count = mean(friend_count),
                                 median_friend_count = median(friend_count),
                                 n = n())

# removes grouping (when using group_by 2 or more variables, operations will be constrained to groups, we need to remove this constrain)
pf.fc_by_age_gender <- ungroup(pf.fc_by_age_gender)

pf.fc_by_age_gender <- arrange(pf.fc_by_age_gender,age)

head(pf.fc_by_age_gender, 10)
```

An alternative way to obtain the same dataframe is using ```%>%```

```{r}

pf.fc_by_age_gender <- pf %>%
                        filter(!is.na(gender)) %>%
                        group_by(age, gender) %>%
                        summarise(mean_friend_count = mean(friend_count),
                                  median_friend_count = median(friend_count),
                                  n = n()) %>%
                        ungroup() %>%
                        arrange(age)
        

head(pf.fc_by_age_gender, 10)

```


***

### Plotting Conditional Summaries

After we consolidated our dataframe we can use it to plot the median of friend count per gender and age.

```{r Plotting Conditional Summaries}
ggplot(aes(x = age, y = median_friend_count), data = pf.fc_by_age_gender) +
  geom_line(aes(color = gender) )
```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Let's reshape median_friend_count data so that the gender is split in a wide view of the data, meaning that gender column will be left out and we will have columns female and male. We can do this using ```reshape2``` library.

```{r}
install.packages('reshape2')
library(reshape2)

pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')

head( pf.fc_by_age_gender.wide,10)
```

Analternative way of reshaping our data is by using ```tidyr```:

```
pf.fc_by_age_gender.wide <-
  subset(pf.fc_by_age_gender[c('age', 'gender', 'median_friend_count')],
         !is.na(gender)) %>%
  spread(gender, median_friend_count) %>%
  mutate(ratio = male / female)

head(pf.fc_by_age_gender.wide)
```


***

### Ratio Plot
Notes:

```{r Ratio Plot}

```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}

```

***

### Cut a Variable
Notes:

```{r Cut a Variable}

```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}

```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}

```

***

### Friending Rate
Notes:

```{r Friending Rate}

```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}

```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}

```

***

### Number of Purchases
Notes:

```{r Number of Purchases}

```

***

### Prices over Time
Notes:

```{r Prices over Time}

```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}

```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

