R course Lesson 3 - Pseudo Facebook
========================================================

[Dataset](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/pseudo_facebook.tsv) taken from Udacity R course lesson 3

***

### What to Do First?
Notes: EDA is an opportunity to let data surprise you and learn from it. EDA will help you understand your variables and that will need you to look at data and make some good visualizations. Also, it is very important that you think back about what question you want to answer with your data.

***

### Pseudo-Facebook User Data
Loading the data **pseudo_facebook.tsv** from the directory in which we stored it and using the right delimiter (the extension *\*.tsv* means *tab separated file*).

```{r Pseudo-Facebook User Data}
# Check current path
path <- getwd()

# Set path to dataset
path <- setwd('~/github/DataAnalyst/Rbasics_Appendix/pseudoFacebookAnalysis/')

file <- 'pseudo_facebook.tsv'

# Read file
pf <- read.csv(file, sep='\t') 
# can use read.delim(file) as its default value for separator is \t
```

***

### Histogram of Users' Birthdays
Plot a histogram of the day of birthday of users using the qplot function from ggplot2 library.
>Note: We can change the theme in which we plot the histogram using the ggthemes library. We can set the theme with theme_set() function. Other ways to set themes and more themes available at [ggthemes github](https://github.com/jrnold/ggthemes#geoms).

```{r Histogram of Users\' Birthdays}
# Load ggplot2 library
#install.packages('ggplot2')
library(ggplot2)

# Load ggthemes
#install.packages('ggthemes', dependencies = TRUE)
library(ggthemes)

# Set the theme to theme_minimal and font size 24
theme_set(theme_minimal(8))

# Show the columns of the dataset
names(pf)

# Plot a histogram with default configuration
qplot(data = pf, x = dob_day)

```

After plotting the data we noticed that looking at it is not very easy, so we need to improve the visualization of this histogram. We can show all the days in the x axis using ```scale_x_continuos()``` function.

```{r}
# Plot a histogram with all days of a month shown as x ticks
# qplot(data = pf, x = dob_day) + scale_x_continuous(breaks=1:31)

# Also plotting the same as above but using ggplot instead of qplot function
ggplot(aes(x = dob_day), data = pf) +
  geom_histogram(binwidth = 0.5 ) +
  scale_x_continuous(breaks = 1:31)

```

***

#### What are some things that you notice about this histogram?
Notice that in this plot the first day has a lot more users, which would mean that more people is borned on the first day of a month or that there is a default day of birthday when signing up. On the other hand, we can see that 31 is the lowest bar and it can be caused due to half of the months have only 30 days.

***

### Moira's Investigation 
We are going to reproduce a study done by a data scientist at Facebook, Moira Burke. 

***

### Estimating Your Audience Size
Moira and her team analyzed users publications and audience and found people tend to understimate their audience size. You can find more in their paper about ["Quantifying the Invisible Audience in Social Networks"](http://hci.stanford.edu/publications/2013/invisibleaudience/invisibleaudience.pdf).

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
A trip to Japan

#### How many of your friends do you think saw that post?
Around 200 of them

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
I think only 20% of my friends saw it.

***

### Perceived Audience Size
As Moira and her team found that usually the audience was underestimated by a quarter of the size, we are going to plot histograms to understand what people guessed about their audience size.

***
### Faceting
Before we plotted all our days of the year/month in one single plot. We would like to separate that plot in histograms per month and look at them. We can use faceting to do this. There are two options for faceting, here we will use ```facet_wrap()``` as it is easier to use when dealing with one variable. However, when dealing with more than one variable it might be better to use ```facet_grid()```, to learn more about how to use it read this [documentation](http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/).

```{r Faceting}

# Plot with month as facet (3 columns, meaning 3 months per row)
ggplot(aes(x = dob_day), data = pf) +
      geom_histogram(binwidth = 0.5) +
      scale_x_continuous(breaks = 1:31) +
      facet_wrap(~dob_month, ncol = 2)

```

#### Let’s take another look at our plot. What stands out to you here?
Three main observations:
* January first is the highest bar and maybe it is not normal as all other days in the year are not even a quarter of it. 
* Some months wihtout 31 day on them have no birthdays which is a good indication 
* February 29 has the smallest amount a day can have birthdays on and that can be related to the calendar in which only every 4 years we have February 29.

***

### Be Skeptical - Outliers and Anomalies
The approach to deal with outliers depends on what your analysis is looking to answer since outliers can have many causes, for example:

* Accurate data about an extreme case
* Bad data
* Limitations of data

One of the approaches used is to replace extreme values with normal values or mean values.

***

### Moira's Outlier
In this section, we will be checking an outlier in the data of the audience estimation analysis.

#### Which case do you think applies to Moira’s outlier?
Bad data of an extreme case since you cannot have 1,000,000 friends on facebook and the data showed 10,000,000 friends watched a post. 
***

### Friend Count


#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(data = pf, x = friend_count)
```

#### How is this plot similar to Moira's first plot?
It has some values that are very large and do not let us appreciate all the data properly.

***

### Limiting the Axes
In order for us to better see and understand the data, we want to limit the axes. We have two options:
* ```xlim``` parameter inside ```qplot()```
* ```scale_x_continuous(limits = c(low,high))`` function added to ```qplot()```

```{r Limiting the Axes}

# option 1, use xlim
qplot(data = pf, x = friend_count, xlim = c(0,1000))

#option 2, use x_scale_continuous
qplot(data = pf, x = friend_count) + scale_x_continuous(limits=c(0,1000))
```

### Exploring with Bin Width
In the plot above, we can see that we have bins in ranges, it will be good to change the ```binwidth``` to explore our data and find how it all distributes with different bin ranges. It will be a good idea to show more ticks by using the ```breaks``` argument in ```scale_x_continuous()```.

***

### Adjusting the Bin Width
```{r Adjusting Bins}
qplot(data = pf, x = friend_count, binwidth=25) +
  scale_x_continuous(limits = c(0,1000), breaks = seq(0,1000,50))
```

### Faceting Friend Count
Now, it will be interesting to split the histogram by gender to understand what males or females think about their audience.
```{r Faceting Friend Count}
# Facet the histogram by gender
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)
```

***

### Omitting NA Values
We noticed that the plot above had a third plot with title NA. This means some users did not define their gender. To remove the NA data we can ```subset``` our dataset using ```!is.na()``` with the ```gender``` column.

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}

```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}

```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}

```

***

### Labeling Plots
Notes:

```{r Labeling Plots}

```

***

### User Ages
Notes:

```{r User Ages}

```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}

```

***


### Frequency Polygons

```{r Frequency Polygons}

```

***

### Likes on the Web
Notes:

```{r Likes on the Web}

```


***

### Box Plots
Notes:

```{r Box Plots}

```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}

```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}

```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}

```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}

```

Response:

***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!