---
title: "diamondsManyVarEDA"
author: "Arturo Parrales"
date: "7/22/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Diamond Dataset Exploration for Many Variables

In this project we will analyze the diamonds dataset from the ggplot package. It is intended to analyze multiple variables with different plots.

```{r diamond}
#Load ggplot first
library(scales)
library(ggplot2) 

# Load diamonds dataset contained in ggplot package
data(diamonds)
d <- diamonds

```

Once we loaded the dataset, we can see the diamonds dataset has:

```{r dataset size, echo=FALSE}
print('Observations: ')
nrow(d) 
print('Variables: ')
ncol(d)
```

## Exploring the Diamonds Price

```{r Histogram of Prices}
# Create a histogram of diamond prices.
# Facet the histogram by diamond color
# and use cut to color the histogram bars.

ggplot(aes(price, fill = cut), data = subset(d, !is.na(price))) +
  facet_wrap( ~ color) +
  geom_histogram() +
  scale_x_log10() +
  scale_fill_brewer(type = 'qual')

```

```{r Scatterplot Price vs Table}
# Create a scatterplot of diamond price vs.
# table and color the points by the cut of
# the diamond.

ggplot(aes(y = price, x = table), data = subset(d, !is.na(table), !is.na(price))) +
  geom_jitter(alpha = 1/3, aes(color = cut)) +
  scale_colour_brewer(type = 'qual') +
  scale_x_continuous(limits = c(50,80), breaks = seq(min(d$table), max(d$table), by = 4))

```

```{r Scatterplot Price vs Volume}
# Create a scatterplot of diamond price vs.
# volume (x * y * z) and color the points by
# the clarity of diamonds. Use scale on the y-axis
# to take the log10 of price. You should also
# omit the top 1% of diamond volumes from the plot.

d <- transform(d, volume = x*y*z)

ggplot(aes(x = volume, y = price), data = subset(d, d$volume <= quantile(d$volume, 0.99))) +
  geom_jitter(aes(color = clarity)) +
  scale_y_log10() +
  scale_color_brewer(type = 'div')

```

```{r Scatterplot Price/Carat Ratio vs Cut}
# Create a scatter plot of the price/carat ratio
# of diamonds. The variable x should be
# assigned to cut. The points should be colored
# by diamond color, and the plot should be
# faceted by clarity.

ggplot(aes(x = cut, y = price/carat), data = subset(d, !is.na(price), !is.na(carat))) +
  geom_jitter(aes(color = color), size = 0.2) +
  facet_wrap( ~ clarity ) +
  ylab('Price/Carat') +
  xlab('Cut') +
  theme(axis.text.x = element_text(face="bold", size=6, angle=90),
        axis.text.y = element_text(face="bold", size=8, angle=0),
        strip.text = element_text(size=10), 
        legend.text = element_text(size=6), legend.title = element_text(face='bold',size=8),
        axis.title.x = element_text(size=12), axis.title.y = element_text(size=12)) +
  scale_colour_brewer(type = 'div') 
  
```

## Predictions on Diamond Price

Now, we will use linear regression models to predict the proce of diamonds. Let's start by creating a scatter plot of price (y) vs carat weight (x)

```{r Scatterplot Price vs Carat}
# Let's consider the price of a diamond and it's carat weight to create a scatterplot

ggplot(aes(x = carat, y = price), data = d) +
  geom_jitter(aes(color = color)) +
  scale_colour_brewer(type = 'div') +
  xlim(0, quantile(d$carat, 0.99)) +
  ylim(0, quantile(d$price, 0.99)) +
  stat_smooth(method='lm', linetype = 2, color = 'black')

```

We can see that the relation is not perfectly linear. Thus, we will if the price is linearly correlated to another variable.

```{r Install and Load Packages }
# Install necessary packages (if not installed before)
#install.packages('GGally')
#install.packages('scales')
#install.packages('memisc')
#install.packages('lattice')
#install.packages('MASS')
#install.packages('car')
#install.packages('reshape')
#install.packages('plyr')

# Load packages
library(ggplot2)
library(GGally)
#library(scales)
library(memisc)

```

```{r ggpairs 1}

set.seed(20022012)
d_samp <- d[sample(1:length(d$price), 10000), ]

ggpairs(d_samp, 
        lower = list(continuous = wrap("points", shape = I('.'))), 
        upper = list(continuous = wrap("points", shape = I('.'))),
        axisLabels = 'internal')

```

Once we have these plots, we can see that there is no linear relationship of price and other variables. We can see there are exponential.

Lets try to understand the price using histograms

```{r Price Histograms}
# The first plot is a histogram of price
# and the second plot transforms
# the price variable using log10.
#install.packages('gridExtra')
library(gridExtra)

plot1 <- qplot(price, data = d, geom = 'histogram', binwidth = 100) + 
  ggtitle('Price')

plot2 <- qplot(price, data = d, geom = 'histogram', binwidth = 0.01) +
  scale_x_log10() +
  ggtitle('Price (log10)')

grid.arrange(plot1, plot2, ncol=2)

```

Going back a bit to the price and carat scatterplot, we can try to use a log scale to see how it changes the plot.

```{r Scatterplot Price vs Carat Log}
# Let's consider the price of a diamond and it's carat weight to create a scatterplot

ggplot(aes(x = carat, y = price), data = d) +
  geom_jitter(aes(color = color)) +
  scale_colour_brewer(type = 'div') +
  xlim(0, quantile(d$carat, 0.99)) +
  ylim(0, quantile(d$price, 0.99)) +
  scale_y_log10() +
  ggtitle('Log10(Price) by Carat')

```

The data looks less disperse. It seems the log function might have helped. However, we might want to use also a cube root function. Let's create a cube root function in R.

```{r Cube Function}
# cube root function
cuberoot_trans = function() trans_new('cuberoot',
                                      transform = function(x) x^(1/3),
                                      inverse = function(x) x^3 )

```

Let's transform the carat value with the cube root function

```{r Cuberoot Carat}
# Transform Carat with cuberoot
ggplot(aes(x = carat, y = price), data = d) +
  geom_point(aes(color = color), alpha=1/2, size = 3/4, position = 'jitter') + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2,3),
                     breaks = c(0.2,0.5,1,2,3)) +
  scale_y_continuous(trans = log10_trans(), limits = c(350,15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  scale_colour_brewer(type = 'div', 
                      guide = guide_legend(title = 'Color', reverse = FALSE,
                                          override.aes = list(alpha = 1, size = 2))) +
  ggtitle('Log10(Price) vs Carat^(1/3)')

```

Now, the relationship between price and carat seems almost linear with the transformations done: 
Log10(Price) and (carat)^(1/3).

We also used the color of the diamond to color our plot and we notice that the color of a diamond also seems related to the price of diamonds. 

**Note:** D is the best color, and J is the worst color of a diamond

Let's explore the same plot, but using clarity as the color of the plot.

```{r Log10(price) vs (carat)^(1/3) and color Clarity}
# analyze the clarity of a diamond, the price and carat
library(RColorBrewer)

ggplot(aes(x = carat, y = price), data = d) + 
  geom_point(aes(color = clarity), alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Clarity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
    breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
    breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Clarity')
```

It seems clarity helps explains the range of prices of diamonds with similar carat.

Let’s look at the diamond cut and see if we find a similar result.

```{r Log10(Price) vs (Carat)^(1/3) and color Cut}
# Color the points by cut

ggplot(aes(x = carat, y = price), data = d) +
  geom_point(aes(color = cut), alpha = 1/2, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Cut', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(trans = cuberoot_trans(),
                     limits = c(0.2, 3), breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) + 
  ggtitle('Price (log10) by Cube-Root of Carat and Cut')
  
```

It seems cut is not useful to define the price of a diamond based on the carat of it. The colors are spread and they do not seem to follow a trend. There are more green colors, but they are in all ranges. The only one that seems to show a pattern is the Ideal cut which seems to be related to more expensive diamonds.

Once we have identified some characteristics on our relation price vs carat, we can start to build a linear model using the log10(price) and (carat)^(1/3)

## Linear Model

```{r Linear Model Log10(Price) vs (Carat)^(1/3)}
# Linear model to predict a diamond price

m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = d)
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)

mtable(m1, m2, m3, m4)

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
