---
title: "Diamonds.rmd"
author: "Arturo Parrales"
date: "5/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Diamond Dataset Exploration

In this project we will analyze the diamonds dataset from the ggplot package.

```{r diamond}
#Load ggplot first
library(ggplot2) 

# Load diamonds dataset contained in ggplot package
data(diamonds)

```

Once we loaded the dataset, we can see the diamonds dataset has:

```{r dataset size, echo=FALSE}
print('Observations: ')
nrow(diamonds) 
print('Variables: ')
ncol(diamonds)
```

After loading the dataset, we can learn more from its documentation by typing ```?diamonds```. Somehting to notice about diamonds:
* Color of diamond has scale from J (worst) to D (best)
* Clarity of a diamond is measured on how clear a diamond is. The scale goes from I1 (worst), SI1, SI2, VS1, VS2, VVS1, VVS2, IF (best).

We can look at the datatypes in our dataset using ```str```

```{r}
str(diamonds)
```

From our datatypes, we can see that we have 3 ordered factor variables: cut, color and clarity.

If we are more curious about our dataset, we can make a summary of our dataset.

```{r diamonds summary}
summary(diamonds)
```

***

## Distribution of Price

Now, we can try to understand the distribution of the price of diamonds

```{r}
summary(subset(diamonds$price, !is.na(diamonds$price)))

qplot(x = price, data = subset(diamonds, !is.na(price)), 
  binwidth = 500,
  color = I('black'), fill = I('#099DD9'),
  xlab = 'Price',
  ylab = 'Number of Diamonds') +
  scale_x_continuous(breaks = seq(0,20000,1250))
```

From our summary statistics and histogram we can tell that the distribution of price is long-tailed and right-skewed. The median is 2401, while the mean is 3933. This confirms that the plot is correct as the mean is greater than the median. Thus, for this type of distribution, our median is a better measure of center.

## Price Subsets

Now, we want to answer:

How many diamonds cost less than $500 ?

We can start doing logical operations in our dataset

```{r}
d_500 <- ifelse(diamonds$price < 500, 1, 0)
sum(d_500)
```

We have that 1729 diamonds cost less than $500.

What about diamonds that cost less than $250 ?
```{r}
d_250 <- ifelse(diamonds$price < 250, 1, 0)
sum(d_250)
```

There are no diamonds that cost less than $250.

If we look at the other side of the spectrum, we would like to know if there are much more expensive diamonds equal or greater than $15000.

```{r}
d_15000 <- ifelse(diamonds$price >= 15000, 1, 0)
sum(d_15000)
```

Our answer is yes, there are very expensive diamonds that wort $15000 or more. 

All these values we got match what we believed:

The minimum value of the price data is \$386 from our summary statistics.Therefore, we expected 0 counts of diamonds smaller than \$250.

The maximum value of the price is \$18823. Thus, we expected to see a value greater than 0 for diamonds of price \$15000 or greater since the distribution histogram shows some diamonds are far on the right of the price distribution tail.


## Exploration of Histogram's Peak

Now, that we explored our ranges on the price data. Let's analyze our peak value in the distribution.

We can zoom in to the range of the highest peaks in the distribution. This range goes from our lowest value 386 to 1250. In order to see our full range and the values next to them, we will choose a range from 300 to 1500. Then, we can actually move our binwidth until we find a value that allows to judge the plot.

```{r}
qplot(x = price, data = subset(diamonds, !is.na(price)), 
  binwidth = 100,
  color = I('black'), fill = I('#099DD9'),
  xlab = 'Price',
  ylab = 'Number of Diamonds') +
  coord_cartesian(xlim = c(300, 1500)) +
  scale_x_continuous(breaks = seq(0,1500,100))
```

After zooming in the peaks, we chose a bindwidth of size 100 and we were able to clearly see that between 550 and 850 is where we find the greater number of diamonds. The highest range is 650, 750. 

We can also see that there are almost no diamonds sold from the range 386 to 400, and 1450 to 1500.

However, it will be great to know the exact price of the highest number of diamonds. We will need to zoom in even further in our range.

```{r}
qplot(x = price, data = subset(diamonds, !is.na(price)), 
  binwidth = 15,
  color = I('black'), fill = I('#099DD9'),
  xlab = 'Price',
  ylab = 'Number of Diamonds') +
  coord_cartesian(xlim = c(550, 850)) +
  scale_x_continuous(breaks = seq(0,850,15))
```

After zooming in more, we didn't get a decent distribution, but we can find a finer max range in the histogram and it goes from \$696.5 to \$712.5. This narrows down our value and we can continue narrowing it down until we find a specific value, but we decided to say the range is good enough.

What we can say about diamonds is that once we found the highest peak it actually means that the cheapest diamonds are more common and among those cheapest diamonds, the ones on the range of \$695.5, \$712.5 are the most common diamonds to find. So if you are expecting to buy a diamond, you might have a good chance to get one as long as you carry \$720 on your pocket.


## Diamond Cut and Price

At this point we might want to find the cut distributions related to the price of a diamond. We know that we have 5 types of cuts, so we are going to plot 5 histograms.

```{r cut and price}
qplot(x = price, data = subset(diamonds, !is.na(price)),
      binwidth = 1000,
      color = I('black'), fill = I('#099DD9'),
    xlab = 'Price',
    ylab = 'Number of Diamonds') +
  facet_wrap(~cut, ncol = 2, scales = 'free_y')
```

We can see that the distributions look quiet similar, except for the fair cut which seems closer to normal. However, all distributions are right-skewed and long-tailed including fair cut. Their peaks also are on the left side, meaning that the lowest priced diamonds are more common for all cuts. Although, fair and good cut diamonds are rarer than all other cut diamonds, and fair cut being the rarest of all.

We can perform summary statistics to know more about each distributions

```{r}
by(diamonds$price, diamonds$cut, summary)
```

After our summary statistics we can say that:
The Premium cut has the highest priced diamond.
The Premium and Ideal cut have the lowest priced diamond.
The Ideal cut has the lowest median priced diamond.

Thus, we can find that Premium cut has the highest and lowest priced diamonds, intersting isn't it? It might be that the premium and ideal cut are making cheaper diamonds rise their price, but not enough to pass the lowest priced diamonds with Fair, Good and Very Good cut.

## Price per Carat and Cut

Once we got familiar with the price and cut distributions, we can think of the carat of a diamond and the price together. It might be good to get the price per carat distribution and then separate such distribution on the different cut types.

```{r}
qplot(x = price/carat, data = subset(diamonds, !is.na(price)),
      binwidth = 300 ,
      color = I('black'), fill = I('#099DD9'),
      xlab = 'Price per Carat',
      ylab = 'Number of Diamonds')
```

Our distribution of price per carat shows that we are right-skewed. We are close to be a normal distribution. We should try the log10 and see if we actually get a normal distribution.

```{r}
qplot(x = price/carat, data = subset(diamonds, !is.na(price)),
      binwidth = 0.1 ,
      color = I('black'), fill = I('#099DD9'),
      xlab = 'Log Scale Price per Carat',
      ylab = 'Number of Diamonds') +
  scale_x_log10()
```

After applying the log10 we got normal distribution. Now, it is time to split this normal distribution in each cut.

```{r}
qplot(x = price/carat, data = subset(diamonds, !is.na(price)),
      binwidth = 0.08 ,
      color = I('black'), fill = I('#099DD9'),
      xlab = 'Log Scale Price per Carat',
      ylab = 'Number of Diamonds') +
  scale_x_log10() +
  facet_wrap(~cut, ncol = 2, scales = 'free_y')
```

After splitting the dsitribution in cut types, we noticed that the Very Good cut has a slight bimodal distribution while other cuts are very clode to normal distributions.

Let's look at these distributions using boxplots to confirm what we saw in the histograms.

```{r}
qplot(x = cut , y = price/carat, 
      data = subset(diamonds, !is.na(price)),
      geom='boxplot') +
  scale_y_log10()
```

After looking at the box plots, we can tell that the distributions seem normal, including the Very Good cut. However, we know that the Very Good cut has a slight bimodal distribution. We can also confirm that the Fair cut is the rarest cut on diamonds and also the most normal distributions.

## Price and Color

After we found some characteristics of price and cuts, we will now check price and color of diamonds.

```{r}
by(diamonds$price, diamonds$color, summary)

qplot(x = color, y = price, data = subset(diamonds,!is.na(price)),
      geom = 'boxplot')
```

Something interesting about diamonds can be seen in the boxplot. The best diamonds color D has more outliers than most of the other colors and its distribution shows that on average it has the second lowest price. For the color E it turns out it has the most outliers and the lowest average price. In fact, color D and E [1st, 3rd] quartiles and IQR are:
D: [911, 4214], IQR 3303
E: [882, 4003], IQR 3121

While for J (worst color) the [1st, 3rd] quartiles and IQR are [1860, 7695] and 5835, respectively.

## 

##

```{r pressure, echo=FALSE}
plot(pressure)
```

**Note:** The `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
